Title

### 2024-06-21 paper/main.tex
Updates

A lot of questions in life sciences center around a causal understanding
between variables. \emph{Causal Discovery} aims to find these casual structure
among random variables from a given dataset. The problem of causal discovery
has been studied both the in Directed Acyclic Graph (DAGs) and Structure
Equation Models (SEM) literatures. In the field of DAGs, many automated
algorithms have been developed to learn the casual structure form a given 
dataset such as constraint-based methods, score-based methods, and continuous
optimization based methods.

@@

However, the adaption of these algorithms in applied fields remain extremely
limited. In these applied fields, even though these algorithms exist,
researchers typically prefer to construct these DAGs manually based on their
domain knowledge. We hypothesize that this preference towards manual DAG
construction stems from the following difficulties in applying causal discovery
methods:

\begin{enumerate}
	\item Difficult to trust: Although asypotically these methods are 
		consistent, in the finite sample scenario they can end up
		making obvious mistakes. As a result of this, researchers
		have to rely on posthoc modification of the output of these
		algorithms using either their domain knowledge or assisted
		by testing methods.
	\item Algorithms output the Markov Equivalence Class (MEC): With any
		causal discovery algorithm, we can only get to the MEC. However,
		most of the methods developed for downstream tasks such 
		as identification or casual effect estimation assume the
		knowledge of a DAG. To orient these MECs to a DAG, researchers
		have to either rely on their domain expertise, or make further
		assumption to utilize a pairwise orientation method.
	\item Difficulties in choosing an algorithm: All the algorithms
		have the same goal with all of them having similar asympottic
		properties. However, the finite sample properties of these
		methods vary depending on the choices that we make. For
		example, the choice of the conditional independence (CI) test
		being used for constraint-based method depends on the dataset,
		similarly for the scoring method for score-based methods. With
		no obvious way to test the impact/performance of these choices
		on a given dataset, it becomes quite hard to chose the optimal
		algorithm for a given dataset.
\end{enumerate}

@@

In the case of DAGs, global testing methodology does exist such as Shipley
test, that can combine local structure tests into a single p-value test.
However, this does not give us a way to make modifications to our model. A
possible way to do this by checking which of these local tests are violated.
However, this approach can quickly become diffult as the size of the network
increases. For example, the popular alarm network implies $x$ condition independence
statements.

One of the drawbacks of these global fit metrics is that they do not only test
for the structure but also the fitted parameterization.

We present a similar procedure for modifying DAGs based on expert knowledge and
measure of association. We first use the measure of association to test whether
the association observed in the data is explained by the model structure. We
can not directly use the local CI tests to make modifications because that
would give way too many independencies and we have no way to rank them. Based
on this testing, we suggest adding edges by ranking them by the most
unexplained association between variables. The researcher can then use expert
knowledge to decide the direction of the edge between the variables.

Our main contributions in this papers are as follows:
\begin{enumerate}
	\item We present a measure of (conditional) assocation between mixed variables using
		canonical correlations. This mixed variable association is a generalization 
		of some of the other association measures such as partial correlation to 
		mixed variables.
	\item We present a procedure to use this measure of association to
		iteratively modify DAGs using expert knowledge such that it
		improves the explained association between variables.
	\item Lastly, to assist researchers in manully constructing and modifying DAGs,
		we developed a web-based tool to iteratively modify their DAGs.
\end{enumerate}

@@

\section{Background}
We represent uni-dimensional variables as $ X $ and multi-dimensional variables
as $ \bm{Z} $. We consider the case of estimating association between $ X $,
and $ Y $ is the presence of a set of conditioning variables $ \bm{Z} = Z_1,
\cdots, Z_k $, potentially being an empty set $ \bm{Z} = \emptyset $. We use $
\rvert \bm{Z} \rvert $ to denote the cardinality of the variable $ \bm{Z} $. We
consider various scenarios where the variables $ X $, $ Y $ and $ \bm{Z} $
could be either discrete, ordinal, or continuous. With mixed data, we mean an
arbitrary mix of variable types among $ X $, $ Y $, and $ \bm{Z} $.

@@

\section{A Measure of Association for Mixed Data}
In the previous section, we outlined measure of associations for different data
types, however we do not have any measure of association for a mixture of
categorical, ordinal, and continuous data. In this section, we develop a
measure of association for mixed data as a generalization of partial
correlation by combining a mixed data residualization approach with canonical
correlations. Similar to partial correlation, we first compute the residuals of
both variables using some machine learning model, and then use a canoncial
correlation based measure of association between the residuals instead of
correlation coefficient.

@@

\section{Using measure of association for modifying DAGs}
The idea behind using the measure of association for modifying DAGs is that,
for any given model and a dataset, we can use the condtiional measure of
association to compute the amount of correlation in the data that is still not
explained by the model. Based on this, we can compute how much improvement in
the explanation would happen if we were to add an edge between variables. This
can then be ranked using the maximum improvement. Give some examples here.

@@

\section{Web Tool}
Based on the approach outlined in the previous section, we built a web-based
tool that researchers can upload their dataset to and apply this method to
interactively draw models.

@@

\section{Empirical Analysis}

>>>>

Many questions in science center around a causal understanding between
variables. \emph{Causal Discovery} aims to find the underlying causal structure
among random variables from a given dataset. The problem of causal discovery
has been studied both in Directed Acyclic Graphs (DAGs) and Structural Equation
Models (SEM) literatures. In the DAG literature, the focus has been on
developing automated algorithms that can learn the causal structure from a
given observational or interventional dataset, whereas in the SEM literature
the focus has on developing tools to help researchers construct or modify
manually constructed models.

@@

In the DAG literature, many automated algorithms have been developed for causal
discovery such as constraint based method (PC, FCI), score-based methods
(Hill-Climb Search, Greedy Equivalence Search), and optimization based methods.
However, the adaption of these algorithms in applied fields remain limited. In
these applied fields, researchers typically prefer to construct these DAGs
manually based on their domain knowledge. We hypothesize that this preference
towards manual DAG construction stems from the following difficulties in
applying causal discovery methods:

\begin{enumerate}
	\item \textbf{Difficult to trust: } Even though most of these algorithms
		have asymptotic theory proving them to be consistent, their
		finite sample properties is not well understood. In real
		setting with finite samples, these algorithms can make obvious
		mistakes. As a result of this, researchers have to rely on
		post-hoc modification of the output of these algorithms
		using either their domain knowledge or assisted by model testing
		methods.
	\item \textbf{Algorithms output Markov Equivalence Class (MEC):} Given an 
		observational dataset, multiple causal structures can be
		faithful to it. As a result, using only observational datasets,
		these algorithms can only get to the MEC such as Completed
		Partially Directed Acyclic Graph (CPDAG) or a Partial Ancestral
		Graph (PAG) where not all the edges are directed. However, as
		most of the methods developed for downstream tasks such as
		identification or casual effect estimation assume the knowledge
		of a DAG. To orient these MECs to a DAG, researchers have to
		either rely on their domain expertise, or make further
		assumption to utilize a pairwise orientation method.
	\item \textbf{Difficult to choose appropriate algorithm and
		hyperparameters:} All the algorithms have the same goal with
		all of them having similar asymptotic properties. However, the
		finite sample properties of these methods vary depending on the
		choices that we make. For example, the choice of the
		conditional independence (CI) test being used for
		constraint-based method depends on the dataset, similarly for
		the scoring method for score-based methods. With no obvious way
		to test the impact/performance of these choices on a given
		dataset, it becomes quite hard to chose the optimal algorithm
		for a given dataset.
\end{enumerate}

@@

As it is evident that researchers also draw DAGs by hand and even if they use
an algorithm, the need to make some post-hoc adjustments to the model, in this
paper we present a tool to assist researchers in making these adjustments
somewhat inspired by modification indices. In the case of DAGs, global testing
methodology does exist such as Shipley test, that can combine local structure
tests into a single p-value test. However, this does not give us a way to make
modifications to our model. A possible way to do this by checking which of
these local tests are violated. However, this approach can quickly become
diffult as the size of the network increases. For example, the popular alarm
network implies $x$ condition independence statements.

% One of the drawbacks of these global fit metrics is that they do not only test
% for the structure but also the fitted parameterization.

Our proposed approach uses a measure of association to find correlated variables
in the data that are not explained by the current model. Using the measure, we 
can rank the edge that can lead to the highest explanation of the observed 
correlation. Based on this information, the user can select a potential edge
and based on their expert knowledge decide the direction of the edge.

% We present a similar procedure for modifying DAGs based on expert knowledge and
% measure of association. We first use the measure of association to test whether
% the association observed in the data is explained by the model structure. We
% can not directly use the local CI tests to make modifications because that
% would give way too many independencies and we have no way to rank them. Based
% on this testing, we suggest adding edges by ranking them by the most
% unexplained association between variables. The researcher can then use expert
% knowledge to decide the direction of the edge between the variables.

Our main contributions in this papers are as follows:
\begin{enumerate}
	\item We present a measure of (conditional) association for mixed data
		based on canonical correlations. This measure of association
		generalizes the widely used partial correlation to mixed data
		setting.
	\item Using this association measure, we present a procedure to suggest
		potential new edges, or edge removals in the model. The user
		gets to decide the direction of the new edge based on their
		expertise.
	\item Lastly, we present a web-browser based tool to allow researchers
		to apply this method to their own datasets.
\end{enumerate}

@@

The rest of the paper is structured as follows. In
Section~\ref{sec:background}, we give a background on the measures of
association for various data types. In Section~\ref{sec:mixed_association}, we
present our method for computing mixed data association. In
Section~\ref{sec:modification}, we present how to use this method to rank the
modifications, and lastly in Section~\ref{sec:empirical}, we show some
empirical analysis on how well this modification process works.

\section{Background}
\label{sec:background}
We represent uni-dimensional variables as $ X $ and multi-dimensional variables
as $ \bm{Z} $. We consider the case of estimating association between $ X $,
and $ Y $ is the presence of a set of conditioning variables $ \bm{Z} = Z_1,
\cdots, Z_k $, potentially being an empty set $ \bm{Z} = \emptyset $. We use $
\rvert \bm{Z} \rvert $ to denote the cardinality of the variable $ \bm{Z} $. We
consider various scenarios where the variables $ X $, $ Y $ and $ \bm{Z} $
could be either discrete, ordinal, or continuous. With mixed data, we mean an
arbitrary mix of variable types among $ X $, $ Y $, and $ \bm{Z} $.

@@

\section{A Measure of Association for Mixed Data}
\label{sec:mixed_association}
In the previous section, we outlined measure of associations for different data
types, however we do not have any measure of association for a mixture of
categorical, ordinal, and continuous data. In this section, we develop a
measure of association for mixed data as a generalization of partial
correlation by combining a mixed data residualization approach with canonical
correlations. Similar to partial correlation, we first compute the residuals of
both variables using some machine learning model, and then use a canoncial
correlation based measure of association between the residuals instead of
correlation coefficient.

@@

\section{Using measure of association for modifying DAGs}
\label{sec:modification}
The idea behind using the measure of association for modifying DAGs is that,
for any given model and a dataset, we can use the condtiional measure of
association to compute the amount of correlation in the data that is still not
explained by the model. Based on this, we can compute how much improvement in
the explanation would happen if we were to add an edge between variables. This
can then be ranked using the maximum improvement. Give some examples here.

@@

\section{Web Tool}
\label{sec:web}
Based on the approach outlined in the previous section, we built a web-based
tool that researchers can upload their dataset to and apply this method to
interactively draw models.

@@

\section{Empirical Analysis}
\label{sec:empirical}
